The Way Forward in Parallelizing Dynamic Languages

### 2014-05-02 talk/icooolps2014/position-paper.tex
start citing things

\usepackage[utf8]{inputenc}
\usepackage{array}
\usepackage{amsmath}

@@

Jython\footnote{www.jython.org} is one project that implements an
interpreter for Python on the JVM\footnote{Java Virtual Machine} and
that uses fine-grained locking to correctly synchronise the
interpreter. For a language like Python, one needs quite a few,
carefully placed locks. Since there is no central location, the
complexity of the implementation is quite a bit greater compared to
using a GIL. Integrating external, non-thread-safe libraries should
however be very simple too. One could simply use one lock per library
to avoid this issue.

@@

HTM in these CPUs works on the level of caches. This has a few
consequences like false-sharing on the cache-line level, and most
importantly it limits the amount of memory that can be accessed within
a transaction. This transaction-length limitation makes it necessary
to have a fallback in place in case this limit is reached. In recent
attempts, the usual fallback is the GIL (XXX: cite). The current
generation of HTM hits this limit very often for our use case (XXX:
cite ruby GIL paper) and therefore does not parallelise that well.

The performance of HTM is pretty good (XXX: cite again...) as it does
not introduce much overhead. And it can transparently parallelise
existing applications to some degree. The implementation is very
straight-forward because it directly replaces the GIL in a central
place. HTM is also directly compatible with any external library that
needs to be integrated and synchronised for use in multiple
threads. The one thing that is missing is support for a better
synchronisation mechanism for the application. It is not possible
in general to expose the hardware-transactions to the application
in the form of atomic blocks because that would require much
longer transactions.

@@

\paragraph{STM} provides all the same benefits as HTM except for its performance.
It is not unusual for the overhead introduced by STM to be between
100\% to even 1000\%. While STM systems often scale very well to a big
number of threads and eventually overtake the single-threaded
execution, they often provide no benefits at all for low numbers of
threads (1-8). There are some attempts (XXX: cite fastlane) that can
reduce the overhead a lot, but also scale very badly so that their
benefit on more than one thread is little.

@@

>>>>

\usepackage[utf8]{inputenc}
\usepackage{array}
\usepackage{hyperref}
\usepackage{amsmath}

@@

Jython\cite{webjython} is one project that implements an
interpreter for Python on the JVM\footnote{Java Virtual Machine} and
that uses fine-grained locking to correctly synchronise the
interpreter. For a language like Python, one needs quite a few,
carefully placed locks. Since there is no central location, the
complexity of the implementation is quite a bit greater compared to
using a GIL. Integrating external, non-thread-safe libraries should
however be very simple too. One could simply use one lock per library
to avoid this issue.

@@

HTM in these CPUs works on the level of caches. This has a few
consequences like false-sharing on the cache-line level, and most
importantly it limits the amount of memory that can be accessed within
a transaction. This transaction-length limitation makes it necessary
to have a fallback in place in case this limit is reached. In recent
attempts, the usual fallback is the GIL\cite{odaira14}. In our
experiments, the current generation of HTM proved to be very fragile
and thus needing the fallback very often. Consequently, scalability
suffered a lot from this.

The performance of HTM is pretty good as it does not introduce much
overhead ($<40\%$ overhead\cite{odaira14}). And it can transparently
parallelise existing applications to some degree. The implementation
is very straight-forward because it directly replaces the GIL in a
central place. HTM is also directly compatible with any external
library that needs to be integrated and synchronised for use in
multiple threads. The one thing that is missing is support for a
better synchronisation mechanism for the application. It is not
possible in general to expose the hardware-transactions to the
application in the form of atomic blocks because that would require
much longer transactions.

@@

\paragraph{STM} provides all the same benefits as HTM except for its
performance.  It is not unusual for the overhead introduced by STM to
be between 100\% to even 1000\% \cite{cascaval08,drago11}. While STM
systems often scale very well to a big number of threads and
eventually overtake the single-threaded execution, they often provide
no benefits at all for low numbers of threads (1-8). There are some
attempts \cite{warmhoff13} that can reduce the overhead a lot, but
also scale very badly so that their benefit on more than one thread is
little.

@@

\bibitem{webjython}
  The Jython Project, \url{www.jython.org}

\bibitem{odaira14}
  Odaira, Rei, Jose G. Castanos, and Hisanobu Tomari.  "Eliminating
  global interpreter locks in Ruby through hardware transactional
  memory."  \emph{Proceedings of the 19th ACM SIGPLAN symposium on
    Principles and practice of parallel programming.} ACM, 2014.

\bibitem{warmhoff13}
  Wamhoff, Jons-Tobias, et al. "FastLane: improving performance of
  software transactional memory for low thread counts."
  \emph{Proceedings of the 18th ACM SIGPLAN symposium on Principles
    and practice of parallel programming.} ACM, 2013.

\bibitem{drago11}
  Dragojević, Aleksandar, et al. "Why STM can be more than a research
  toy." \emph{Communications of the ACM} 54.4 (2011): 70-77.

\bibitem{cascaval08}
  Cascaval, Calin, et al. "Software transactional memory: Why is it
  only a research toy?." \emph{Queue} 6.5 (2008): 40.
